The \tsc{r}-pronoun and postposition \tit{waarmee} in Dutch

### 2020-01-16 paper.tex
bij jan vandaan
intro ver af

English has a sentence. This is ungrammatical in Dutch. We need an \tsc{r}-pronoun. The term \tsc{r}-pronoun \citep{riemsdijk1978} refers to a set of nominal elements that can strand prepositions in Dutch (and German). This is noteworthy because Dutch is a normally non-preposition stranding language.
In the example \ref{ex:klimerop} \tit{op} `on' is the preposition and \tit{'r} `there' is the locative pronoun.\footnote{The \tsc{r}-pronoun \tit{'r} in \ref{ex:klimerop} can be written as \tit{er}, \tit{der} and \tit{'r} and pronounced as respectively /ɛr/, /dər/ or /ər/. As far as I am aware, there is no clear meaning difference between these forms. See \citet{wesseling2018} for discussion. In my examples I use \tit{'r}, but the other two forms fit just as well.}

\ex. I climb on it.

\ex.
\ag. *Ik klim op 't.\\
 I climb on it\\
 `I climb on it.'\label{ex:klimopt}
\bg. Ik klim 'r op.\\
 I climb there on\\
 `I climb on it.'\label{ex:klimerop}

In Dutch it is impossible to have [P inanimate]. But why? \citet{riemsdijk1978} postulating a filter that prohibits the existence of constituents consisting of prepositions and inanimates + some kind of allomoprhy. \citet{koopman2000} who makes reference to a different paradigm. Abels: base-generated somewhere else.

This chapter is an in-depth study of the \tsc{r}-pronoun-postposition combination \tit{waarmee} `with what' in Dutch. This instance is interesting for two reasons. First, just like for all \tsc{r}-pronouns, the wh-element is the locative, but there is no meaning component related to location in `with what'. Second, the preposition \tit{met} `with' does not only turn into a postposition, but it also changes into \tit{-mee} `with' when it is combined with an \tsc{r}-pronoun.\footnote{I do not have anything to say about the distinction between prefixes and prepositions, or suffixes and postpositions.} This last observervation has so far remained unexplained.

In answering the question why [P inanimate] I look at an exception to the rule that *P inanimate pronoun. That exception is a mismatching free relative. Here the string \tit{met wat} does appear.

what I will show that r-pronouns are not a different construction to express the same meaning, but they realize the same features as regular P + pronoun combinations. It is a kind of allomorphy, just like Van Riemsdijk said. And not a different way to express the same function.

In this paper I show that distribution of \tit{waarmee} `with what' and \tit{met wat} `with what' in these free relative constructions gives us a good insight into the internal structure of \tsc{r}-pronouns. In what follows I show that \tsc{r}-pronouns and regular preposition compete to spell out the same syntactic features. If all features form a proper constituent (i.e. a constituent to the exclusion of other features), the \tsc{r}-pronoun surfaces. If it is not a proper constituent, the preposition-pronoun combination shows up. This straightforwardly follows in a system in which spellout targets phrasal constituents: Nanosyntax \citep{starke2009}. --here more about the internal structure of r-pronouns, and the postposition preposition distinction--

This paper is structured as follows. In Section \ref{sec:distribution} I discuss the distribution of \tit{waarmee} `waarmee' and \tit{met wat} `with what'. I show that the \tsc{r}-pronoun surfaces when it forms a proper constituent, and \tit{met wat} `with what' when it does not. I decompose \tit{waarmee} `with what' and \tit{met wat} `with what' to make sense of the phonological similarities, and I provide a more detailed analysis. Unmarked examples are constructed and have been verified by native speakers.

\section{The distribution between \tit{waarmee} and \tit{met wat}}\label{sec:distribution}

@@

The goal of this section is to show that \tit{waarmee} `with what' is the default as instrumental relative pronoun. This generalization is not new, it has already been made \citet{riemsdijk1978,koopman2000}. In order to show that \tit{waarmee} `with what' is the default, I discuss the distribution of \tsc{r}-pronouns and regular pronouns in more general. I start with the personal pronouns and then return to the wh-pronouns.

@@

\ex. \label{ex:objprep}
\ag. Ik schilder samen met haar/hem.\\
 I paint together with her/him\\
 `I am painting together with her/him.'\label{ex:prepani}
\bg. *Ik schilder met 't.\\
 I paint with it\\
 `I am painting with it.'\label{ex:prephet}
\bg. Ik schilder 'r mee.\\
 I paint there-with\\
 `I am painting with it.'\label{ex:preper}
\bg. *Ik schilder mee 'r.\\
 I paint with there\\
 `I am painting with it.'\label{ex:erprep}

@@

\subsection{\tit{Waarmee} is a constituent, \tit{met wat} is not}

@@

In this section I showed that ʀ-pronouns are expected in combinations with prepositions. This means that the use of \tit{waarmee} `with what' \ref{ex:grammaticalwaar} is not surprising. Something that is surprising is the use of \tit{met wat} `with what' in \ref{ex:grammaticalwat}, and this is the example something more needs to be said about. In the remainder of this section I argue that this `something more' is that the instrumental object in \ref{ex:grammaticalwat} does not form a proper constituent, i.e. it is not a constituent to the exclusion of any other elements. The other side of the coin is that constructions with \tsc{r}-pronouns contain a object that does form a proper constituent.

@@

%\ex.\label{ex:constwaarmee}
%\ag. Waar-mee schilder jij?\\
 %{where-with} paint you\\
 %`What are you painting with?'
%\bg. Ik schilder [waar-mee jij ook schildert].\\
 %I paint {where-with} you also paint\\
 %`I am painting with what you are painting with too.'
%\bg. Ik schilder met de kwast [waar-mee jij ook schildert].\\
 %I paint with the brush {where-with} you also paint\\
 %`I am painting with the brush that you are painting with too.'

@@

I put \tit{w} and \tit{a} aside for now, assuming they spell out the same syntactic structure in \tit{waarmee} `with what' and \tit{met wat} `with what'. This leaves \tit{'r-mee} `with it' and \tit{met 't} `with it'.

\ex.
\ag. 'r -mee\\
there with\\
\b. met 't\\
with it\\

@@

In this section I discuss the forms \tit{met 't} `with it` and \tit{'rmee} `with it'. Below I repeat examples from Section \ref{sec:rdefault} that show that \tit{met} `with` can express instrumental case, and \tit{'t} is an inanimate pronoun. \ref{ex:metdp} shows \tit{met} `with' combined with full DP. \ref{ex:tverb} shows \tit{'t} as the object of verbs.

However, \tit{met} `with' and \tit{'t} `it' do not appear together if they form a proper constituent. Instead, \tit{'rmee} is used, as shown in \ref{ex:rmeeconst}.

\ex.\label{ex:rmeeconst}
\ag. Ik schilder 'r-mee.\\
 I paint there-with\\
 `I am painting with it.'\label{ex:jarmee}
\bg. *Ik schilder met 't.\\
 I paint with it\\
 `I am painting with it.'\label{ex:neemett}

In this section I set up an account that makes the ungrammaticality of \tit{met 't} and the appearance of \tit{'rmee} follow from spellout. The analysis accounts for the following two observations, taking \tit{met 't} as the point of departure. First, \tit{met} `with' changes from being a preposition to being a postposition, and its form changes into \tit{-mee}. This process is restricted to inanimate pronouns, and it does not apply to full DPs and animate pronouns. Second, \tit{'t} is replaced by \tit{'r}, a morpheme that is associated with the locative in Dutch.

@@

\begin{table}[ht]
	\center
	\caption {Fragment Dutch \tsc{n.sg}}
	\begin{minipage}{0.56\linewidth}
		\begin{tabularx}{\textwidth}{ccccccc}
		\toprule
              & \tsc{n.sg} \\
		\midrule
    \tsc{nom} & 't         \\
    \tsc{acc} & 't         \\
    \tsc{dat} & 't         \\
    \tsc{gen} & 'r-van     \\
    \tsc{ins} & -r-mee     \\
    \bottomrule
\end{tabularx}
\label{tbl:dutchcases}
\end{minipage}
\end{table}

@@

\ex.
\ag. Ik schilder 'r-mee.\\
 I paint there-with\\
 `I am painting with it.'\label{ex:dutchina}
\bg. *Ik schilder met 't\\
 I paint with it\\
 `I am painting with it.'\label{ex:dutchinmet}

>>>>

Dutch has the preposition \tit{met} `with' that expresses instrumental case. In \ref{ex:metdp}, \tit{met} `with' is combined with a full DP. The inanimate pronoun in Dutch is \tit{'t} `it'.\footnote{The longer form that is mostly used in writing is \tit{het} `it'. I will use the spoken variant \tit{'t} throughout this paper.} \ref{ex:tverb} illustrates that \tit{'t} `it' can be used as the object of a verb.

\tit{Met} `with' and \tit{'t} `it' do not appear together, as illustrated in \ref{ex:neemett}. Instead, Dutch uses the \tsc{r}-pronoun \tit{'r} and the postposition \tit{mee}, as shown in \ref{ex:jarmee}.\footnote{The \tsc{r}-pronoun \tit{'r} in \ref{ex:klimerop} can be written as \tit{er}, \tit{der} and \tit{'r} and pronounced as respectively /ɛr/, /dər/ or /ər/. As far as I am aware, there is no clear meaning difference between these forms. See \citet{wesseling2018} for discussion.
In my examples I use \tit{'r}, but the other two forms fit just as well.}$^,$\footnote{In this paper I do not make any claims about the distinction between prefixes and prepositions, or suffixes and postpositions.}

\ex.\label{ex:rmeeconst}
\ag. *Ik schilder met 't.\\
 I paint with it\\
 `I am painting with it.'\label{ex:neemett}
\bg. Ik schilder 'r -mee.\\
 I paint there -with\\
 `I am painting with it.'\label{ex:jarmee}

\tsc{R}-pronouns \citep{riemsdijk1978,koopman1994} are nominal elements that are syncretic with locative pronouns, which in Dutch means they contain the morpheme \tit{r}. The adposition it combines with obligatorily follows the \tsc{r}-pronoun, compare \ref{ex:jarmee} and \ref{ex:neemeer}.
Notice also that the preposition \tit{met} `with' has changed form into \tit{mee} `with' when it appears as a postposition (see \ref{ex:metdp} and \ref{ex:jarmee}).

\exg. *Ik schilder mee 'r.\\
 I paint with there\\
 `I am painting with it.'\label{ex:neemeer}

This chapter is concerned with the question of how to correctly rule out the construction in \ref{ex:neemett} and to let the \tsc{r}-pronoun construction in \ref{ex:jarmee} appear. I argue that \tsc{r}-pronouns are not something special, but a consequence of regular spellout. Just like \citet{riemsdijk1978} I analyze \tsc{r}-pronouns and postpositions as a type of allomorphy. What is different is that spellout rules out the ungrammaticality of the preposition and inanimate pronoun, rather than the stipulation of a filter.

This chapter is an in-depth study of the instrumental \tsc{r}-pronoun-postposition combination \tit{'r -mee} `with it' and \tit{waarmee} `with what' in Dutch. This instance is interesting for two reasons. First, just like for all \tsc{r}-pronouns, the \tsc{r}-pronoun is syncretic with the locative. In many of the \tsc{r}-pronoun-postpositions combinations, the meaning component of the locative is intuitive, as many prepositions express locations, directions etc. However, an instrumental expresses an instrument and does not have a meaning component assosicated with location.
The second reason why I focus on this particular \tsc{r}-pronoun and postposition has to do with the form of the adposition. The preposition \tit{met} `with' does not only turn into a postposition, but it also changes into \tit{-mee} `with' when it is combined with an \tsc{r}-pronoun.\footnote{I do not have anything to say about the distinction between prefixes and prepositions, or suffixes and postpositions.} This last observervation has so far remained unexplained.

%%here the basic idea%%
When all features form a proper constituent, then the \tsc{r}-pronoun surfaces. When there is a complement or whatever, it does not work.
In what follows I show that \tsc{r}-pronouns and regular preposition compete to spell out the same syntactic features. If all features form a proper constituent (i.e. a constituent to the exclusion of other features), the \tsc{r}-pronoun surfaces. If it is not a proper constituent, the preposition-pronoun combination shows up. This straightforwardly follows in a system in which spellout targets phrasal constituents: Nanosyntax \citep{starke2009}. --here more about the internal structure of r-pronouns, and the postposition preposition distinction--

%%here the structure of the paper%%
This paper is structured as follows. In Section \ref{sec:distribution} I discuss the distribution of \tit{waarmee} `waarmee' and \tit{met wat} `with what'. I show that the \tsc{r}-pronoun surfaces when it forms a proper constituent, and \tit{met wat} `with what' when it does not. I decompose \tit{waarmee} `with what' and \tit{met wat} `with what' to make sense of the phonological similarities, and I provide a more detailed analysis. Unmarked examples are constructed and have been verified by native speakers.

\section{\tsc{R}-pronouns are proper constituents}\label{sec:distribution}

@@

The goal of this section is to show that \tit{waarmee} `with what' is the default as instrumental relative pronoun, a shown in a.o. \citet{riemsdijk1978,koopman2000}. In order to show that \tit{waarmee} `with what' is the default, I discuss the distribution of \tsc{r}-pronouns and regular pronouns in more general. I start with the personal pronouns and then return to the wh-pronouns.

@@

\ex. \label{ex:objprep}
\ag. Ik schilder samen met haar/hem.\\
 I paint together with her/him\\
 `I am painting together with her/him.'\label{ex:prepani}
\bg. *Ik schilder met 't.\\
 I paint with it\\
 `I am painting with it.'\label{ex:prephet}
\bg. Ik schilder 'r mee.\\
 I paint there -with\\
 `I am painting with it.'\label{ex:preper}
\bg. *Ik schilder mee 'r.\\
 I paint with there\\
 `I am painting with it.'\label{ex:erprep}

@@

\subsection{\tit{Met wat} `with what' shows up}

Sometimes \tit{met wat} shows up!

@@

In the preceding section I showed that ʀ-pronouns are expected in combinations with prepositions. This means that the use of \tit{waarmee} `with what' \ref{ex:grammaticalwaar} is not surprising. Something that is surprising is the use of \tit{met wat} `with what' in \ref{ex:grammaticalwat}, and this is the example something more needs to be said about. In the remainder of this section I argue that this `something more' is that the instrumental object in \ref{ex:grammaticalwat} does not form a proper constituent, i.e. it is not a constituent to the exclusion of any other elements. The other side of the coin is that constructions with \tsc{r}-pronouns contain a object that does form a proper constituent.

@@

%\ex.\label{ex:constwaarmee}
%\ag. Waar -mee schilder jij?\\
 %{where -with} paint you\\
 %`What are you painting with?'
%\bg. Ik schilder [waar -mee jij ook schildert].\\
 %I paint {where -with} you also paint\\
 %`I am painting with what you are painting with too.'
%\bg. Ik schilder met de kwast [waar -mee jij ook schildert].\\
 %I paint with the brush {where -with} you also paint\\
 %`I am painting with the brush that you are painting with too.'

@@

I put \tit{w} and \tit{a} aside for now, assuming they spell out the same syntactic structure in \tit{waarmee} `with what' and \tit{met wat} `with what'. This leaves \tit{'r -mee} `with it' and \tit{met 't} `with it'.

\ex.
\ag. 'r -mee\\
there with\\
\b. met 't\\
with it\\
\z.
\z.

@@

In this section I discuss the forms \tit{met 't} `with it` and \tit{'rmee} `with it'. I set up an account that makes the ungrammaticality of \tit{met 't} and the appearance of \tit{'rmee} follow from spellout. The analysis accounts for the following two observations, taking \tit{met 't} as the point of departure. First, \tit{met} `with' changes from being a preposition to being a postposition, and its form changes into \tit{-mee}. This process is restricted to inanimate pronouns, and it does not apply to full DPs and animate pronouns. Second, \tit{'t} is replaced by \tit{'r}, a morpheme that is associated with the locative in Dutch.

@@

\begin{table}[ht]
	\center
	\caption {Fragment Dutch \tsc{n.sg}}
	\begin{minipage}{0.56\linewidth}
		\begin{tabularx}{\textwidth}{ccccccc}
		\toprule
              & \tsc{n.sg} \\
		\midrule
    \tsc{nom} & 't         \\
    \tsc{acc} & 't         \\
    \tsc{dat} & 't         \\
    \tsc{gen} & 'r-van     \\
    \tsc{ins} & -r -mee     \\
    \bottomrule
\end{tabularx}
\label{tbl:dutchcases}
\end{minipage}
\end{table}

@@

\ex.
\ag. Ik schilder 'r -mee.\\
 I paint there -with\\
 `I am painting with it.'\label{ex:dutchina}
\bg. *Ik schilder met 't\\
 I paint with it\\
 `I am painting with it.'\label{ex:dutchinmet}
